
---
title: "Project: Analysis of the effect of alcohol consumption and social background on academic performance"
output: html_notebook
---

**Names**

  - Boshra ARIGUIB 
  - Rogério KACIAVA BOMBARDELLI
  - Juliana CASTELO BRANCO DREYER


*Libraries*
```{r}
library(corrplot)
library(reshape2)
library(ggplot2)
library(tidyverse)
library(viridis)
library(hopkins)
library(patchwork)
library(ggeffects)
library(ROSE)
library(caret)
library(e1071)
library(dplyr)
library(multcomp)
library(pROC)
library(cluster)
library(randomForest)
```

# 1. Presentation of the problem
- Goal: Analyse the effect of alcohol consumption and family background on academic performance
- Importance: help in the decision process or design of school management systems

# 2. Overview of data
## Data Acquisition
```{r}
students <- read.csv("student-lpor.csv", header = TRUE, sep = ",") 
```

## Data Analysis
```{r}
glimpse(students)
```

### Barplots for categorical data
```{r}
for (col in names(students) ) {
  plot <- ggplot(students, aes_string(x = col)) +
          geom_bar(position = position_dodge(), fill = '#be4d25') +
          theme_classic()
    print(plot)
}
```

### Statistical summaries for numerical data
```{r}
summary_data <- summary(students[, sapply(students, is.numeric)])
summary_data
```

# 3. Data preparation
## Check for missing values
```{r}
print("Number of missing values for each row :")
na_total <- lapply(students, function(x) sum(is.na(x)))  
na_total
```
According to the above output, there is not any missing value in the dataset. We can continue next steps.

## Adjust the types of the variables
```{r}
students <- mutate_if(students, is.character, as.factor)
```
We change all the character types to factor 

## Target variable 
The mean of the the grades from the first and second semester turned to levels 'bad' and 'good'
```{r}
students$G_mean <- (students$G1 + students$G2) / 2
students$grade <- ifelse(students$G_mean <= 12, "bad", "good")
students$grade <- factor(students$grade, levels = c('bad', 'good'))
tab <- table1::table1(~ . | grade, students)
tab
```

# 3. Exploratory data analysis

## Correlation Matrix 
*Full matrix*
```{r}
encoded_students <- model.matrix(~ . - 1, students)
cormat <- round(cor(encoded_students),2)
melted_cormat <- melt(cormat)
ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile()
```
*Correlated variables of interest*
```{r}
substudents = students[c('Medu', 'Fedu', 'address', 'traveltime', 'Pstatus', 'famrel')] 
substudents <- mutate_if(substudents, is.character, as.factor)
substudents <- mutate_if(substudents, is.factor, as.integer)
corrplot(cor(substudents), col = colorRampPalette(c("#be4d25", "#9925be", "#2596be"))(100), tl.col = 'black', insig='blank', type = 'lower')#, addCoef.col ='black', number.cex = 0.6)
```
```{r}
substudents = students[c('studytime', 'freetime', 'goout', 'absences')] 
substudents <- mutate_if(substudents, is.character, as.factor)
substudents <- mutate_if(substudents, is.factor, as.integer)
corrplot(cor(substudents), col = colorRampPalette(c("#be4d25", "#9925be", "#2596be"))(100), tl.col = 'black', insig='blank', type = 'lower')#, addCoef.col ='black', number.cex = 0.6)
```
## Distribution of outcome

**ANOVA Analysis**
```{r}
anova_result <- aov(G_mean ~ school + sex + age + address + famsize + Pstatus + Medu + Fedu + traveltime + famrel + studytime + freetime + goout + absences + health + Walc + Dalc, data = students)
print(summary(anova_result))
```
```{r}
cor_result <- cor.test(students$G_mean, students$Dalc)

# Check correlation coefficient and p-value
cor_result

```
```{r}
substudents = students[c('G_mean', 'school', 'sex', 'age', 'address', 'famsize', 'Pstatus', 'Medu', 'Fedu', 'traveltime', 'famrel', 'studytime', 'freetime', 'goout', 'absences', 'health', 'Walc',  'Dalc')] 
substudents <- mutate_if(substudents, is.character, as.factor)
substudents <- mutate_if(substudents, is.factor, as.integer)
corrplot(cor(substudents), col = colorRampPalette(c("#be4d25", "#9925be", "#2596be"))(100), tl.col = 'black', insig='blank', type = 'lower')#, addCoef.col ='black', number.cex = 0.6)

```

**As Barplots**
```{r}
# Plot grade levels by absense levels 

for (col in c('sex', 'school', 'address')) {
plot <- ggplot(students, aes_string(x = col, fill = 'grade')) +
    geom_bar(position = "dodge") +
    scale_fill_manual(values = c("#2596be", "#9925be", "#be4d25", "#49be25", "#102444")) +
    theme_classic()
print(plot)
}
```

**As Boxplots**
```{r}
# Define variable names for the plot
var_names <- list(sex = "Sex", school = "School", famrel = "Family Relation", health = "Health status")

for (col in c('famrel', 'health')) {
  plot <- ggplot(data = students, 
           aes_string(x = paste('factor(', col, ')') , y = 'G_mean', group = paste('factor(', col, ')'), fill = paste('factor(', col, ')'))) +
           labs(y = 'Average Grade', x = var_names[make.names(col)]) +
           scale_fill_manual(values = c("#2596be", "#9925be", "#be4d25", "#49be25", "#1054a4")) + #102444
           geom_boxplot(show.legend = FALSE) +
           ggtitle(glue::glue("Distribution of Grades per {var_names[make.names(col)]}"))
  print(plot)
}
```

**As Lineplots**
```{r}
avg_grade_by_dalc_wd <- aggregate( students$G_mean ~ students$Dalc, data = students, mean)
avg_grade_by_walc_wd <- aggregate( students$G_mean ~ students$Walc, data = students, mean)
plot(avg_grade_by_dalc_wd, type = 'b', col='#9925be', xlab = "Average alcohol consumption", ylab = "Grade")
lines(avg_grade_by_walc_wd, type = 'b', col='#2596be')
legend("topright", legend = c("Weekdays", "Weekends"), col = c("#9925be", "#2596be"), lwd = 2)
```


## 4. Model Learning

### 1. Class balance & Test-Train datasets
Check distribution of the variable stroke in the data
```{r}
round(prop.table(table(students$grade)), 2)
```

To balance the classes, we oversample the minority class using ROSE 
```{r}
set.seed(15)
students_rose <- ROSE(grade ~ ., data = students)$data
round(prop.table(table(students_rose$grade)), 2)
```

### 2. Classification
Classification Task: 
- Given: a set of features (all variables in the dataset or a subset)
- Output: predict is the student gets a good or bad grade

### SVM
```{r}
set.seed(15)
students_svm = students_rose[c('Walc', 'Dalc', 'grade')]
index <- sample(nrow(students_svm), nrow(students_svm)*0.8)
svm_train <- students_svm[index, ]
svm_test <-  students_svm[-index, ]

# Feature Scaling 
#svm_train[-3] = scale(svm_train[-3]) 
#svm_test[-3] = scale(svm_test[-3]) 
```

```{r}
model_svm_alc = svm(grade ~ ., data = svm_train, kernel = "linear", type = 'C-classification')
plot(model_svm_alc, svm_train, col = c("#2596be", "#9925be", "#49be25", "#be4d25", "#1054a4"))
```
```{r}
trainControl <- trainControl(method="repeatedcv", number=10, repeats=3, verboseIter = TRUE)
model_svm_alc_tuned = svm(grade ~ ., data = svm_train, kernel = "linear", type = 'C-classification', tunecontrol = trainControl)
plot(model_svm_alc_tuned, svm_train, col = c("#2596be", "#9925be", "#49be25", "#be4d25", "#1054a4"))
```
**SVM predictions**
```{r}
# Predictions
predictions <- predict(model_svm_alc, svm_test, type = "response")

#Confusion Matrix
confusionMatrix <- table(predictions, svm_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```
**SVM tuned predictions**
```{r}
# Predictions
predictions <- predict(model_svm_alc_tuned, svm_test, type = "response")

#Confusion Matrix
confusionMatrix <- table(predictions, svm_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```

#### Logisitc Regression
```{r}
set.seed(15)
index <- sample(nrow(students_rose), nrow(students_rose)*0.8)
students_train <- students_rose[index, ]
students_test <-  students_rose[-index, ]

# 10-fold cross validation with 3 repeats
trainControl <- trainControl(method="repeatedcv", number=10, repeats=3, verboseIter = TRUE)
metric <- "Accuracy"

```
We fit an initial logistic regression model to predict the occurrence of a stroke, to find the most significant variables.
**Model**
```{r}
formula <- grade ~ school + sex + age + address + famsize + Pstatus + Medu + Fedu + traveltime + famrel + studytime + freetime + goout + absences + health + Walc + Dalc
model_glm <- glm(formula, data = students_train, family = binomial)
summary(model_glm)
```
**Tuned model**
```{r}
model_glm_tuned <- train(formula, 
                        students_train, method = "glmStepAIC", 
                        metric = "accuracy", 
                        tuneLength = 10,
                        trControl = trainControl)
summary(model_glm_tuned$finalModel) 

```

#### SVM
**SVM Model**
```{r}
model_svm = svm(formula, data = students_train, kernel = "linear")
summary(model_svm)
```
**Tuned SVM Model**
```{r}
model_svm_tuned = svm(formula, data = students_train, kernel = "polynomial", gamma = 0.1, C = 1,  type = 'C-classification', tunecontrol = trainControl)# = tune.control(sampling = "cross", cross = 10), cost = 10)
summary(model_svm_tuned)
```

#### Random Forest
**Random Forest Model**
```{r}
set.seed(15)
model_rf <- randomForest(formula, data=students_train) 
```
**Tuned Random Forest Model**
```{r}
set.seed(15)
model_rf_tuned <- train(formula, 
                        students_train, method = "rf", 
                        metric = "accuracy", 
                        trControl = trainControl)
print(model_rf_tuned)
```

## k-Nearest Neighbour
**kNN**
```{r}
model_knn <- train(formula, students_train, method = "knn", tuneGrid = data.frame(k = c(3)))
summary(model_knn) 
```
**Tuned kNN**
```{r}
model_knn_tuned <- train(formula, students_train, method = "knn", metric = "Accuracy", tuneLength = 10,trControl = trainControl)
summary(model_knn_tuned$finalModel) 
```

#### Prediction
**Prediction GLM**
```{r}
# Predictions
predictions_glm <- predict(model_glm, students_test, type = "response")
predictionsBinary <- ifelse(predictions_glm > 0.5, "good", "bad")

#Confusion Matrix
confusionMatrix <- table(predictionsBinary, students_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```

**Prediction GLM tuned**
```{r}
# Predictions
predictions_glm_tuned <- predict(model, students_test, type="prob")$good
predictionsBinary <- ifelse(predictions_glm_tuned > 0.5, "good", "bad")

#Confusion Matrix
confusionMatrix <- table(predictionsBinary, students_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```

**SVM predictions**
```{r}
# Predictions
predictions_svm <- predict(model_svm, students_test, type = "response")

#Confusion Matrix
confusionMatrix <- table(predictions_svm, students_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```

**SVM tuned predictions**
```{r}
# Predictions
predictions_svm_tuned <- predict(model_svm_tuned, students_test, type = "response")

#Confusion Matrix
confusionMatrix <- table(predictions_svm_tuned, students_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```

**Random Forest**
```{r}
# Predictions
predictions_rf_raw <- predict(model_rf, students_test, type="prob")[, 2]
predictions_rf <- ifelse(predictions_rf_raw > 0.5, "good", "bad")
  
#Confusion Matrix
confusionMatrix <- table(predictions_rf, students_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```
**Tuned Random Forest**
```{r}
# Predictions
predictions_rf_tuned_raw <- predict(model_rf_tuned, students_test, type="prob")$good
predictions_rf_tuned <- ifelse(predictions_rf_tuned_raw > 0.5, "good", "bad")

#Confusion Matrix
confusionMatrix <- table(predictionsBinary, students_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```
**k-Nearest-Neighboor**
```{r}
# Predictions
predictions_knn_raw <- predict(model_knn, students_test, type="prob")[, 2]
predictions_knn <- ifelse(predictions_knn_raw > 0.5, "good", "bad")

#Confusion Matrix
confusionMatrix <- table(predictions_knn, students_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```
**Tuned k-Nearest-Neighboor**
```{r}
# Predictions
predictions_knn_tuned_raw <- predict(model_knn_tuned, students_test, type="prob")[, 2]
predictions_knn_tuned <- ifelse(predictions_knn_tuned_raw > 0.5, "good", "bad")

#Confusion Matrix
confusionMatrix <- table(predictions_knn_tuned, students_test$grade)
print(confusionMatrix)

# Accuracy
accuracy <- sum(diag(confusionMatrix)) / sum(confusionMatrix)
print(paste("Accuracy", accuracy))
```

## RESULTS 

**ROC Curve**
```{r}
roc_glm <- roc(students_test$grade, predictions_glm)
roc_glm_tuned <- roc(students_test$grade, predictions_glm_tuned)
roc_rf <- roc(students_test$grade, predictions_rf_raw)
roc_rf_tuned <- roc(students_test$grade, predictions_rf_tuned_raw)
roc_knn <- roc(students_test$grade, predictions_knn_raw)
roc_knn_tuned <- roc(students_test$grade, predictions_knn_tuned_raw)
```

```{r}
plot(roc_glm, main = "ROC Curve for GLM Model", col = "#2596be", lwd = 2)
lines(roc_glm_tuned,col = "#9925be", lwd = 2)
lines(roc_knn,col = "#be4d25", lwd = 2)
lines(roc_knn_tuned,col = "#1054a4", lwd = 2)
lines(roc_rf,col = "#49be25", lwd = 2)
lines(roc_rf_tuned,col = "#ff64a4", lwd = 2)
#c("#2596be", "#9925be", "#49be25", "#be4d25", "#1054a4"))
# Add diagonal line (random classifier)
abline(a = 0, b = 1, lty = 2, col = "grey")
legend("bottomright", legend = c("GLM Model", "GLM Tuned", "kNN", "kNN Tuned", "Random Forest", "Random Forest Tuned", "Random Classifier"), col = c("#2596be", "#9925be", "#be4d25", "#1054a4", "#49be25", "#ff64a4", "grey"), lty = c(1, 1, 1, 1, 1, 1, 2), lwd = 2)
```

### 3. Clustering
```{r}
#substudents = students[c('G_mean', 'school', 'sex', 'age', 'address', 'famsize', 'Pstatus', 'Medu', 'Fedu', 'traveltime', 'famrel', 'studytime', 'freetime', 'goout', 'absences', 'health', 'Walc',  'Dalc')] 
substudents = students[c('G_mean', 'studytime', 'Dalc')] 
substudents <- mutate_if(substudents, is.character, as.factor)
substudents <- mutate_if(substudents, is.factor, as.integer)
```

**Hopkins score for cluster potential**
```{r}
# check data is suitable for clustering or not  with such as hopkins 
hopkins_statistic <- hopkins::hopkins(substudents)
cat("Hopkins statistic score for the first dataset: ", hopkins_statistic)
```
**K-Means for different values of k**
```{r}
# Run K-Means clustering with different values of k
k_values <- 2:10
silhouette_scores <- sapply(k_values, function(k) {
  kmeans_result <- kmeans(substudents, centers = k, nstart = 25)
  cluster_silhouette <- silhouette(kmeans_result$cluster, dist(df_h1))
  mean(cluster_silhouette[, 3])
})

# Plot Silhouette Score vs. Number of Clusters
plot(k_values, silhouette_scores, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of Clusters (k)", ylab = "Silhouette Score",
     main = "Silhouette Score for Different Numbers of Clusters")
```
**K-means**
```{r}
# Perform k-means clustering (for example, with k = 3)
k <- 5
kmeans_result <- kmeans(substudents, centers = k)

# Plot the data points with colors representing the cluster assignments
plot(substudents, col = kmeans_result$cluster, pch = 16, main = "K-means Clustering")
# Add cluster centers
points(kmeans_result$centers, col = 1:k, pch = 3, cex = 2)
# Add legend
legend("topright", legend = paste("Cluster", 1:k), col = 1:k, pch = 16, cex = 1.2, title = "Cluster")

library(plot3D)
scatter3D(substudents$Dalc, substudents$studytime, substudents$G_mean, colvar = kmeans_result$cluster, pch = 16, cex = 2, col = c("#2596be", "#9925be", "#49be25", "#be4d25", "#1054a4"), theta = 30, phi = 20, ticktype = "detailed", main = "K-means Clustering", xlab="Alcohol", ylab="Studytime", zlab="grades")
```